---
title: "OmicsON"
author: "Monika Piwowar, Wiktor Jurkowski, Ewa Gajda, Cezary Turek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OmicsON}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r global_options, include=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = TRUE)
```

# Table of Contents
i. [About OmicsON](#about-omicson)
i. [OmicsON workflow](#omicson-workflow)
  * 1. [Set up mapping files](set-up-mapping-files)
  * 2. [Data input](#input-data)
  * 3. [Decorate data by Reactome](#uecorate-data-by-reactome)
  * 4. [Expand gene clusters with STRING DB](#use-string-db)
  * 5. [Define group](#define-goups)
  * 6. [Lalalala lala](#define-groups-of-genes-and-metabolites)
  * 7. [Multivariate Statistical Analysis](#statistical-analysis)
    * a. [CCA - Canonical Correlation Analysis](#cca---canonical-correlation-analysis)
    * b. [PLS - Partial Least Squares Regression](#pls---partial-least-squares-regression)
ii. [Methods in deatils](#methods-in-deatils)
  * 1. [clasterUsingOntology()](#clasterusingontology())


## About OmicsON
OmicsON provides knowedge driven data regularisation to facilitate multivariate analysis of 'omics' data.

## OmicsON Workflow
Below you will find OmicsON workflow described step by step. By following these steps and using examplary data included in this package user should get the same results as shown in this vignette. 


### Set up mapping files
It is important to set up OmicsON mapping. You can set it up by invoking function presented below. ChEBI to Reactome, Ensembl to Reactome and UniProt to Reactome mappings are required.

```{r, results = 'asis'}
OmicsON::setUpReactomeMapping(ChEBI2ReactomeFileURL = "https://reactome.org/download/current/ChEBI2Reactome.txt",
                              Ensembl2ReactomeFileURL = "https://reactome.org/download/current/Ensembl2Reactome.txt",
                              UniProt2ReactomeFileURL = "https://reactome.org/download/current/UniProt2Reactome.txt")
```


### Data input
After setting up mapping, to start work with OmicsON you need to provide two 'omics' data sets in data frame form. DF can be created from files as in \extdata directory under package insallation directory. This files are in tab delimited files with headers (colnames and rownames). Files are named `nm-transcriptomics.txt` and `nm-lipidomics.txt`.

Below you can find first few lines of files presented in form of data frame. As you can see all of them have heades and colnames. To find them localization please run `find.package("OmicsON")` after package loading `library(OmicsON)`. You can use below snippet of code to load mentioned files into R environment:

```{r, results = 'asis'}
    pathToFileWithLipidomicsData <- system.file(package="OmicsON", "extdata", "nm-lipidomics.txt")
    lipidomicsInputData <- read.table(pathToFileWithLipidomicsData, header = TRUE)
    lipidomicsInputDf <- head(lipidomicsInputData, 6)
    knitr::kable(lipidomicsInputDf[1:7], caption = "Lipidomisc data")
    
    pathToFileWithTranscriptomicsData <- system.file(package="OmicsON", "extdata", "nm-transcriptomics.txt")
    transcriptomicsInputData <- read.table(pathToFileWithTranscriptomicsData, header = TRUE)
    transcriptomicsInputDf <- head(transcriptomicsInputData, 6)
    knitr::kable(transcriptomicsInputDf[1:7], caption = "Transcriptomics data")
```


### Decorate data by Reactome

The first step is to load the input file. When this is done then we are ready to map ChEBI identifiers contained in the file to Reactome database by searching of ontologically related molecules present in Reactome's pathways. The output includes six rows of mapped ids.

```{r, echo=TRUE, results='asis'}
    decoratedByReactome <- OmicsON::decorateByReactomeData(chebiMoleculesDf = lipidomicsInputData[c(2, 3, 6),],
                                chebiIdsColumnName = "ChEBI", organismTaxonomyId = '9606')
    knitr::kable(decoratedByReactome)
```


The result table represent mapping of all small molecules to respective parents and children of ChEBI ontology: "root" denotes source IDs; if ID already exists in Reactome ontologyId column has the same value, if not but there is an alternative in the form of child or parent in ChEBI ontology, OmicsON put its id under ontologyId column. If we can not find root id and can not find any parent or children for it, OmicsON leave it empty.

Results include ontologyId column which contains chebi ids used in following procedure, Root column show from which point we start ontology search. In whoWins you can find R, C, P, N letters which mean Root, Child, Parent, None.

Once the result table of merge operation is generated, you can use it to find genes which are in the same pathway as small molecules with specific ChEBI identifiers. Currently the only supported source of canonical pathways is Reactome database.  
Additionally instead of using dataframe from merge result (mergedSmallMolecules) you can use any dataframe, but then you have to specify a name of the column included ontology IDs:

As a result you recive dataframe presented below:

Result of `OmicsON::mapReactomePathwaysUnderOrganism` is DF, but this DF is complicated, which include lists as cells. This DF contains respectively:

- root - ChEBIs ids given by user,
- ontologyId - ChEBI ids used in the calculation, it is taken from ChEBI ontology base on root,
- ensembleIds - List including vector of Ensemble's Ids,
- uniProtIds - List including vector of UniProt's Ids,
- ensembleIds - List including vector of Ensemble's Ids,
- reactomeIds - List including vector of pathway's ids from Reactome DB,
- genesSymbolsFromEnsemble - List including vector of gen's symbols from Reactome DB base on pathway and Ensemble's Ids,
- genesSymbolsFromUniProt - List including vector of gen's symbols from Reactome DB base on pathway and UniProt's Ids,

```{r, echo=FALSE, results='asis'}
    oneRowDf <- chebiIdsToReactomePathways[6,]
    rownames(oneRowDf) <- NULL
    knitr::kable(oneRowDf)
```


### Use STRING DB

When you have results from Reactome step, then you are ready to use STRING DB step. In this part you search for any extra interactions of gens which you find in Reactome. STRING calls them neighbours. To do it just put results achived from Reactome to `OmicsON::getStringNeighbours()` method. This function produce dataframe. Below one row of this this dataframe is presented:

```{r, echo=TRUE, results='hide'}
chebiIdsToReactomePathwaysAndToStringNeighbours <- OmicsON::getStringNeighbours(
    chebiIdsToReactomePathways[chebiIdsToReactomePathways$ontologyId == "CHEBI:15756",],
    stringOrganismId = 9606,
    stringDbVersion = "10",
    idsColumnName = 'ontologyId',
    rootColumnName = NULL,
    listOfEnsembleIdColumnName = 'ensembleIds')
```

DF returned from this method is also a complicated one. It includes the same columns as previous and two extra columns. All of theme are from STRING database and has some edges (connections) with gens from ensembleIds column:

- stringIds -
- stringGensSymbols - 


```{r, echo=FALSE, results='asis'}
    chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$stringIds[[1]] <- chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$stringIds[[1]][1:50]
    chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$stringGenesSymbols[[1]] <-
    chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$stringGenesSymbols[[1]][1:45]
    chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$ensembleIds[[1]] <-
    chebiIdsToReactomePathwaysAndToStringNeighbours[chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]$ensembleIds[[1]][1:11]
    knitr::kable(
        chebiIdsToReactomePathwaysAndToStringNeighbours[
            chebiIdsToReactomePathwaysAndToStringNeighbours$ontologyId == "CHEBI:15756",]
    )
```


### Define groups of genes and metabolites
This step joins files with data to analyse with results of Reactome step and/or results of STRING step. 


### Define goups
Groups are defined in data frame. OmicsON has method to create DF from GMT file, so you can use `readGroupsAsDf()` to do it. Example file of grouping is under example directory. 

```{r, echo=TRUE, results='asis'}
    gmtGroupsFilePath <- paste(find.package("OmicsON"),"/example/nm-groups.txt", sep = "")
    groups <- OmicsON::readGroupsAsDf(pathToFileWithGroupDefinition = gmtGroupsFilePath)
```

```{r, echo=FALSE, results='asis'}
    knitr::kable(groups)
```


####User selected genes and metabolites
Optionally, the analysis could be executed by hand picked selections of genes and metabolites

```{r, echo=TRUE, results='asis'}
#select small molecules
lip1 <- mergedSmallMolecules[mergedSmallMolecules$root == "CHEBI:27432",]$root
lip2 <- mergedSmallMolecules[mergedSmallMolecules$root == "CHEBI:73705",]$root
joinLip <- c(as.character(lip1), as.character(lip2))

#use Reactome genes mapped to selected small molecules
reactomeTrans1 <- chebiIdsToReactomePathways[chebiIdsToReactomePathways$ontologyId == "CHEBI:15756",]$genesSymbolsFromEnsemble[[1]]
reactomeTrans2 <- chebiIdsToReactomePathways[chebiIdsToReactomePathways$ontologyId == "CHEBI:16015",]$genesSymbolsFromEnsemble[[1]]
joinRecatomeTrans <- c(reactomeTrans1, reactomeTrans2)[!duplicated(c(reactomeTrans1, reactomeTrans2))]
```

### Functional Ineractions DF
You can create a functional interactions data frame by using this method:

```{r, echo=TRUE, results='asis'}
    functionalInteractions <- OmicsON::createFunctionalInteractionsDataFrame(chebiIdsToReactomePathways)
```

```{r, echo=FALSE, results='asis'}
    knitr::kable(head(functionalInteractions, 6))
```

### Statistical Analysis

OmicsON provides two statistical methods to analyse those data:

- CCA - Canonical Correlation Analysis
- PLS - Partial Least Squares Regression

#### CCA - Canonical Correlation Analysis

Calculate CCA by hand:


```{r, echo=TRUE, results='asis'}
    pathToExampleFileWithXData <- paste(find.package("OmicsON"),"/example/nm-transcriptomics.txt", sep = "")
    pathToExampleFileWithYData <- paste(find.package("OmicsON"),"/example/nm-lipidomics.txt", sep = "")

    XDF <- read.table(pathToExampleFileWithXData, header = TRUE);
    YDF <- read.table(pathToExampleFileWithYData, header = TRUE);

    ccaResults1 <- OmicsON::makeCanonicalCorrelationAnalysis(
        xNamesVector = joinRecatomeTrans,
        yNamesVector = joinLip,
            XDataFrame = XDF,
            YDataFrame = YDF)
```


```{r, fig.show='hold', fig.width=6, fig.height=6}
    OmicsON::plotCanonicalCorrelationAnalysisResults(ccaResults = ccaResults1)
```

Calculate CCA on groups:

```{r, echo=TRUE, results='hide'}
    mccReactome <- OmicsON::makeCCAOnGroups(
        groupsDefinitionDF = groups, 
        mappingDF = chebiIdsToReactomePathwaysWithRoot, 
        groupsDataDF = YDF, 
        mappingDataDF = XDF)

``` 

Inside the makeCCAOnGroups functions the permutatuin test is automaticly done. To do it yourself on your data do:

```{r, echo=FALSE, results='hide'}
    permutationTestsResults <- OmicsON::makePermutationTestOnCCA(
        XDataFrame = XDF, 
        YDataFrame = YDF, 17, 2, 50, 
        countedCCA = ccaResults1)
```

To take data from grouped CCA you should know a little the structure, it is like that:
```{r, echo=FALSE, results='hide'}
mccReactome$Molecules[1]
mccReactome$right[[1]]
mccReactome$left[[1]]
mccReactome$ccaResults[[1]]
mccReactome$ccaPermutationTestResults[[1]]
```

```{r, fig.show='hold', fig.width=6, fig.height=6}
    OmicsON::plotCanonicalCorrelationAnalysisResults(ccaResults = mccReactome$ccaResults[[1]])
```
#### PLS - Partial Least Squares Regression

```{r, echo=FALSE, results='hide'}
    PLSResult1 <- OmicsON::makePartialLeastSquaresRegression(
        joinRecatomeTrans,
        joinLip,
        XDataFrame = XDF,
        YDataFrame = YDF)
```

```{r, fig.show='hold', fig.width=6, fig.height=6}
    OmicsON::plotRmsepForPLS(PLSResult1$training)
```  

```{r, fig.show='hold', fig.width=6, fig.height=6}
    OmicsON::plotRegression(PLSResult1$training, ncompValue = 10)
```

The same but on groups defined by user.
```{r, echo=TRUE, results='asis'}
    groupPlsReactome <- OmicsON::makePLSOnGroups(
        groupsDefinitionDF = groups, 
        mappingDF = chebiIdsToReactomePathwaysWithRoot, 
        groupsDataDF = YDF, 
        mappingDataDF = XDF)

```

How to take counted values from results:
```{r, echo=FALSE, results='hide'}
groupPlsReactome$Molecules[1]
groupPlsReactome$right[[1]]
groupPlsReactome$left[[1]]
groupPlsReactome$plsResults[[1]]
groupPlsReactome$plsPermutationTestResults[[1]]
```
How to use it to plot:
```{r, fig.show='hold', fig.width=6, fig.height=6, echo=TRUE}
    OmicsON::plotRmsepForPLS(groupPlsReactome$plsResults[[1]]$training)
```
```{r, fig.show='hold', fig.width=6, fig.height=6, echo=TRUE}
    OmicsON::plotRegression(groupPlsReactome$plsResults[[1]]$training, ncompValue = 10)
```



